---
title: "Untitled"
author: "Yuqi Zhang"
date: "2025-04-10"
output: html_document
---

**Purpose** To understand interval breast cancer phenotype and prognosis, through timing since last screening and breast density

**Population** 
1. large cohort of patients with detection mode known (BCSTO). Specifically, the patients with their **first BC diagnosis as IntCa or ScrCa** were included, also named **cohort1** in this project. The IntCa or ScrCa were between 1989 Jan to 2020 Mar, and the outcome of death and other cancers were until end of 2023.

Based on cohort1, **cohort1_ic** was defined using filter(outcome_IC==1) for the study on solely IC over timing.

2. small cohort of patients with additional mammography info (**Libro1 + Karma**), also named **cohort2** in this project.

cohort2 is patients diagnosed until 2023.10 combined from two cohorts, with duplicates between two cohorts removed. Data contain basic cols that exist in both, with the exception of calcs and masses, which only exist in karma.

Use filter(cohort == "karma"/"libro1") to select cohort

Use filter(sc_ic == 1) to select IntCa

```{r}
library(tidyverse)
library(haven)
```

```{r setup, include=FALSE}
## load previously prepared dataset
load("/Users/yuqzha/LIBRO1_BCSTO/Stockholm women/Data/Data_output/workdata.Rdata")
cohort1 <- workdata |> 
  filter(outcome_IC==1 | outcome_SDC==1) |> 
  mutate(
         ben = ifelse(ben == "3", "insitu", "invasive"),
         ben = factor(ben, levels = c("invasive", "insitu")),
         histology = case_when(
           snomedo10_c %in% c(85003, 85233) ~ "ductal",
           snomedo10_c == 85203 ~ "lobular",
          #snomedo10_c == 85223 ~ "duc+lob" ## N=0 in the current data
          TRUE ~ "Other"
         ),
         histology = ifelse(
           !is.na(snomedo10_c), histology, NA
         ),
         HER2_positive = case_when(
           HER2_negative == 0 ~ 1,
           HER2_negative == 1 ~ 0,
           TRUE ~ NA_real_
         ),
    age_cat = case_when(
           age_at_diagnosis>=40 & age_at_diagnosis < 50 ~ "40-49",
           age_at_diagnosis>=50 & age_at_diagnosis < 60 ~ "50-59",
           age_at_diagnosis>=60 & age_at_diagnosis < 75 ~ "60-74"
         ),
    month = as.factor(floor(diff_days / 30)), # months since screening
    year_since_screen = ifelse(is.na(diff_days), NA, # diff_days is the interval
                                    ifelse(diff_days <= 365, 0, 1)),
    year_since_screen = factor(year_since_screen),
    year = as.numeric(year(diadat))
    ) |> 
  relocate(HER2_positive, .before = grade) |> 
  relocate(ki67, .after = grade) |> 
  relocate(histology, .after =ki67)

cohort1_ic <- cohort1 |> 
  filter(outcome_IC==1)
```
## FH, more details
### cohort1
```{r}
load("/Users/yuqzha/LIBRO1_BCSTO/Stockholm women/Data/Data_output/FH_HBOC_age.Rdata")

## if var already exist
cohort1 <- cohort1 |> select(!FH_BC_bl_median:FH_hboc_in_bl_40)

cohort1 <- cohort1 |> 
  left_join(FH_hboc_age, by = "id")

## cols FH_BC_bl_median:FH_hboc_in_bl_40 means the number of relative below certain age, and 0 means 0 rel below median age but they are diagnosed
## transformation: group 2 or more together
##                 compute so 0=no FH, and 1=FH, no early-onset, 2=1 or more early-onset (0 is when n_sr>=1, otherwise NA)
library(tidyverse)

# Transform the specified columns using tidyverse functions
cohort1 <- cohort1 %>%
  mutate(across(
    # Select columns from FH_BC_bl_median to FH_hboc_in_bl_40
    .cols = FH_BC_bl_median:FH_hboc_in_bl_40,
    .fns = function(x) {
      # 1. Group numbers larger than 1 into 1
      x <- if_else(x > 1, 1, x, missing = x)
      
      # 2. Add 1 to all values
      x <- x + 1
      
      # 3. Replace NA with 0 when n_p_sib >= 1, otherwise keep as NA
      x <- if_else(!is.na(n_1sr) & n_1sr >= 1, 
                  if_else(is.na(x), 0, x),
                  NA)
      x <- factor(x)
      return(x)
    }
  ))
```


## education
```{r}
library(haven)
otherfactors <- read_sas("/Users/yuqzha/LIBRO1_BCSTO/Data/fromQiaoli/for_yuqizhang_allvar.sas7bdat")

cohort1 <- cohort1 |> 
  left_join(otherfactors, join_by(id == studieid)) 
  
cohort1 <- cohort1 |>
  mutate(Education = factor(Education),
         age_1st_child = case_when(age_child_min<30 & age_child_min>0 ~ "<30",
                                   age_child_min>=30 ~ ">=30",
                                   age_child_min==0 ~ "None"),
         age_1st_child = factor(age_1st_child, levels = c("None","<30", ">=30")))
```

## treatments
```{r}
setwd("/Users/yuqzha/LIBRO1_BCSTO/Stockholm women/Data/Quality cancer registers")
rcc <- read_sas("rccsthlm_index.sas7bdat", .name_repair = tolower) |> 
  rename(id = starts_with("studie"))

nkbc <- read_sas("nkbc_index.sas7bdat", encoding = "UTF-8") |> 
  rename(id = starts_with("studie"))

names(nkbc) <- tolower(names(nkbc))
```

```{r}
# data required, rcc, nkbc, cohort (id, diadat)
cohort <- cohort1
source("0 DATA treatment.R")
save(trt, file = "/Users/yuqzha/LIBRO1_BCSTO/Stockholm women/Data/Data_output/treatment.RData")

# output cohort_trt: trt with diadats within 6 months and the closest 
names(cohort_trt) # diadat.x in the cohort, diadat.y in trt data
mean(cohort_trt$diadat.x == cohort_trt$diadat.y) # 97% are on exact same date

cohort1 <- cohort1 |> 
  left_join(cohort_trt[,c("id", "post_chemo", "post_radio", "post_endo")])
```

# Cohort2
```{r}
getwd()
load("/Users/yuqzha/LIBRO1_BCSTO/Karma_Libro1/Data_output/karma.Rdata")
load("/Users/yuqzha/LIBRO1_BCSTO/Karma_Libro1/Data_output/libro1.Rdata")

```

## Karma patients: pat
replace Karma baseline mammography with closest to diagnosis (defined in mam_closest_diagnose_cases.R (output: karma_mam_closest_diagnose.Rdata))


```{r}
## update Karma closer to diagnosis info: mamm features, tc
load("/Users/yuqzha/LIBRO1_BCSTO/Karma_Libro1/Data_output/karma_mam_closest_diagnose.Rdata")
load("/Users/yuqzha/LIBRO1_BCSTO/Karma_Libro1/Data_output/Karma_tc_rcc_nkbc_other.Rdata")

## choose patient with bc_diagdate
pat <- karma |> 
  filter(!is.na(bc_diagdate)) # & bc_invasive==1)   ## bc_diagdate is the most up-to-date until 2023.10.27, the date based on nkbc and medrec

## add mam closest to diagnosis
pat <- pat |>
  select(!pd_mlo:masses_bi) |>  # remove mammogram info from baseline
  left_join(karma_mam_closest_diagnose, by = "id") 

## add tc
pat <- pat |> 
  left_join(karma_tc_rcc_nkbc_other, by = c("id", "bc_diagdate")) 
```

## Unify variable names (mainly mammographic features different between karma and libro1)

Karma only
"pd_mlo"  "mammography_date"    "days_mam_diag"            
"masses"  "calcs"  "cad_date"   "days_cad_diag"                       

Libro1 only
[273] "pd_mult_sources"                 "pd_mult_sources_date"  
"postmenopausal"                  "postmenopausal_before_bc" 

```{r}
pat <- pat |> 
 mutate(pd = pd_mlo, 
        pd_date = mammography_date, 
        interval = diff_days,
        cohort = "karma",
        postmenopausal_before_bc = ifelse(is.na(menopause_status), NA,
                                          ifelse(menopause_status == 3, 1, 0)))

libro1 <- libro1 |> 
 mutate(pd = pd_mult_sources, 
        pd_date = pd_mult_sources_date, 
        cohort = "libro1",
        masses = NA_real_,
        calcs = NA_real_,
        cad_date = NA_Date_) 

libro1 <- libro1 |> 
  filter(is.na(participant_alsoin_karma)) ## remove duplicates
```

## Choose covariates
```{r}
# updated June 18, 2025
covar <- intersect(names(pat), names(libro1)) ## common vars
```

## Combine cases
```{r}
cohort2 <- pat[, covar] |> 
   rbind(libro1[, covar])

cohort2 <- cohort2 |> 
  mutate(diadat = bc_diagdate)
```

## common data
### updated calcs and masses in calcs_masses_2506.R
```{r}
oldcols <- c("cad_date", "calcs", "masses", "mammography_date", "calcs_l_nbr", "calcs_r_nbr", "masses_l_nbr", "masses_r_nbr") # mammography_date is not date for pd, which is pd_date

cohort2 <- cohort2 |> 
  select(!any_of(oldcols)) # any_of only matches cols that exist 


cohort2 <- cohort2 |> 
  left_join(mam_bf_4y[, c("studieid", "calcs","calcs_l_nbr", "calcs_r_nbr", "masses","masses_l_nbr", "masses_r_nbr")], join_by(id == studieid)) # alternatively use mam_bf_2y
```

```{r}
cohort2 <- cohort2 |> 
  mutate(
    calcs_sameside = case_when(
      side_right == 0 ~ calcs_l_nbr,
      side_right == 1 ~ calcs_r_nbr),
    calcs_otherside = case_when(
      side_right == 0 ~ calcs_r_nbr,
      side_right == 1 ~ calcs_l_nbr),
    masses_sameside = case_when(
      side_right == 0 ~ masses_l_nbr,
      side_right == 1 ~ masses_r_nbr),
    masses_otherside = case_when(
      side_right == 0 ~ masses_r_nbr,
      side_right == 1 ~ masses_l_nbr),
    calcs_bi = case_when(
      calcs_sameside == 0 ~ 0, 
      calcs_sameside >= 1 ~ 1),
    calcs_bi = factor(calcs_bi, labels = c("0", ">=1")),
    masses_bi = case_when(
      masses_sameside == 0 ~ 0, 
      masses_sameside >= 1 ~ 1),
    masses_bi = factor(masses_bi, labels = c("0", ">=1")),
    masses_3cat = case_when(
      masses_sameside == 0 ~ 0, 
      masses_sameside == 1 ~ 1,
      masses_sameside >=2 ~ 2),
    masses_3cat = factor(masses_3cat, labels = c("0", "1", ">=2")),
    masses_dif = masses_sameside - masses_otherside,
    masses_dif_4cat = case_when(
      masses_dif < 0 ~ 0, 
      masses_dif == 0 ~ 1,
      masses_dif == 1 ~ 2,
      masses_dif >1 ~ 3),
    masses_dif_4cat = factor(masses_dif_4cat, labels = c("<0", "0","1", ">=2")),
    masses_dif_3cat = case_when(
      masses_dif <= 0 ~ 0, 
      masses_dif == 1 ~ 1,
      masses_dif >1 ~ 2),
    masses_dif_3cat = factor(masses_dif_3cat, labels = c("<=0", "1", ">=2")),
    masses_dif_2cat = case_when(
      masses_dif <= 0 ~ 0, 
      masses_dif >= 1 ~ 1),
    masses_dif_2cat = factor(masses_dif_2cat, labels = c("<=0", ">=1"))
 )
```
for corresponding side use masses_l_nbr masses_r_nbr

### prep var
```{r}
source("2.1 var prep lab.R")

getwd()
```

### GRS and ptv (karma and libro combined data)
```{r}
library(readr)
grs <- read_delim("~/LIBRO1_BCSTO/Karma_Libro1/Data_other/GRS_BC_Yuqi/Libro1KARMA_newGRS_Yuqi.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE)
 
grs <- grs |> filter(!is.na(GRS_BC)) ## 29105 left, 46865 removed

grs <- grs |> filter(QC == "Passed QC") ## 20469 left
length(unique(grs$studiepersonid))

ptv <- read_delim("~/LIBRO1_BCSTO/Karma_Libro1/Data_other/PTV_from_Juan/KARMA_Libro1_PTVs_noDUP_For_Yuqi.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE)

```

```{r}
cohort2 <- cohort2 |> 
  left_join(grs, join_by(id == studiepersonid))

sum(!is.na(cohort2$GRS_BC)) ## 8331 (75%) with grs info,
```

```{r}
cohort2 <- cohort2 |> 
  left_join(ptv, join_by(id == studiepersonid))

sum(!is.na(cohort2$protein_LOF_rare_variants_exonic))
```

## FH HBOC age (added Aug 2 2025)
```{r}
setwd("/Users/yuqzha/LIBRO1_BCSTO/Karma_Libro1/Data_output/swecan relative")
load("FH_hboc_age_karma.RData")
FH_hboc_age_karma <- FH_hboc_age
load("FH_hboc_age_libro1.RData")
FH_hboc_age_libro1 <- FH_hboc_age

FH_hboc_age <- bind_rows(FH_hboc_age_karma, FH_hboc_age_libro1)
FH_hboc_age <- FH_hboc_age |> 
  group_by(id) |> 
  slice(1) |> 
  ungroup()

cohort2 <- cohort2 |> 
  left_join(FH_hboc_age, join_by(id == id))

cohort2 <- cohort2 %>%
  mutate(across(
    # Select columns from FH_BC_bl_median to FH_hboc_in_bl_40
    .cols = FH_BC_bl_median:FH_hboc_in_bl_40,
    .fns = function(x) {
      # 1. Group numbers larger than 1 into 1
      x <- if_else(x > 1, 1, x, missing = x)
      
      # 2. Add 1 to all values
      x <- x + 1
      
      # 3. Replace NA with 0 when n_p_sib >= 1, otherwise keep as NA
      x <- if_else(!is.na(n_1sr) & n_1sr >= 1, 
                  if_else(is.na(x), 0, x),
                  NA)
      x <- factor(x)
      return(x)
    }
  ))
```

## treatments added on Aug 9
```{r}
setwd("/Users/yuqzha/LIBRO1_BCSTO/Karma_Libro1/Data_raw/Libro1")
rcc1 <- read_sas("reg_rcc_sthlm_breastcan_prework.sas7bdat", .name_repair = tolower) |> 
  rename(id = starts_with("studie"))

nkbc1 <- read_sas("2025-02-20/reg_nkbc_prework.sas7bdat", .name_repair = tolower) |> 
  rename(id = starts_with("studie"))

setwd("/Users/yuqzha/LIBRO1_BCSTO/Karma_Libro1/Data_raw/Karma")
rcc2 <- read_sas("reg_rcc_sthlm_breastcan_prework.sas7bdat", .name_repair = tolower) |> 
  rename(id = starts_with("studie"))
nkbc2 <- read_sas("2025-02-20/reg_nkbc_prework.sas7bdat", .name_repair = tolower) |> 
  rename(id = starts_with("studie"))
medrec <- read_sas("2025-02-20/medical_records_prework.sas7bdat", .name_repair = tolower)

rcc <- bind_rows(rcc1, rcc2)
nkbc <- bind_rows(nkbc1,nkbc2)

# data required, rcc, nkbc, cohort (id, diadat)
cohort <- cohort2
source("0 DATA treatment.R")
save(trt, file = "/Users/yuqzha/LIBRO1_BCSTO/Karma_Libro1/Data_output/treatment.RData")
# output cohort_trt: trt with diadats within 6 months and the closest 
names(cohort_trt) # diadat.x in the cohort, diadat.y in trt data
mean(cohort_trt$diadat.x == cohort_trt$diadat.y) # 95.6% are on exact same date

cohort2 <- cohort2 |> 
  left_join(cohort_trt[,c("id", "post_chemo", "post_radio", "post_endo")])
```

```{r}
save(cohort1, cohort1_ic, pat, libro1, cohort2, file = "data/cohorts.RData")
```

```{r}
load("data/cohorts.RData")
```

